Demographic inference using the SFS with \moments and \demes

### 2025-07-31 chapter/paper.tex

Using the split-with-migration model defined above (illustrated in
Figure~\ref{fig:im}A), we simulated data for 30 diploid individuals from both
populations using \msprime \cite{baumdicker2022efficient}. These simulations
consided of 500 regions, each of length 1 Mb, with per-base recombination and
mutation rates of $10^{-8}$. The joint SFS was computed using \tskit's
\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
region and summing across regions to find the total SFS across 500 Mb of data
(Figure~\ref{fig:im}B--D).

@@

The model that we wished to fit is relatively simple but still has many 
parameters ($11$ in the example we present), so we constructed it in stages by 
first fitting simpler one-- and two--population models.
Throughout, we used $u=1.5\times 10^{-8}$ as an estimate of the mutation rate per 
generation and 29 years as the average generation time. When the goal is to infer
a complex demographic history involving multiple populations, we typically start
simple, with subsets of the full data, and sequentially add complexity.
Here, we began by fitting 
one-population models with ancestral and recent size changes for MSL and GBR.
For the MSL model, we allowed three epochs with constant effective size in each 
epoch. We attributed to GBR a sharp contraction $\sim$60 kya followed by 
exponential growth to represent the out-of-Africa bottleneck and subsequent 
rapid expansion. YAML files representing the models and parameters in this 
section can be found in the linked GitHub repository.

@@

We next stitched these one-population models together and refitted the 
resulting demographic model to get a two-population model 
that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
(90 kya, \cite{prufer2017high}) in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
(see, e.g., \cite{sumer2025earliest}) may be reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

@@

The site frequency spectrum is just one of many options for inferring
evolutionary parameters from genetic data. It is straightforward to compute
from data, and there are many methods for predicting the SFS, making it an
appealing framework for demographic inference. As shown here in two examples,
we have incorporated the \demes format for specifying demographic models in
\moments to easily compute the SFS and perform demographic inference, including
computing uncertainties.

\moments is limited in the size of the dataset that it can consider. Sample
sizes can be large (in the hundreds), especially when considering models with
one or few populations, but will encounter memory issues when sample sizes are
extreme or as the number of populations grows. We are limited to a maximum of
five populations existing at any one time, including ancient samples. Other
SFS-based methods, such as \momi \cite{kamm2020efficiently, dilber2024faster},
can include additional populations, typically with smaller sample sizes.

@@

When inferring models with multiple populations, especially four or more, both
the numbers of parameters and the space of plausible model topologies can grow
rapidly. It can therefore become challenging for the optimization routines to
find optima in parameter space, and overparameterization can become a problem.
One hint that models are overparameterized (or models misspecified) is runaway
behavior of parameter values, in which they converge to unreasonably small or
large values (or approach the imposed bounds). Starting with simple
parameterizations and sequentially adding complexity helps to alleviate this
issue. The challenge of large model space is more difficult to address, and
must be addressed through building and testing a wide range of plausible model
topologies. While time-consuming, a broad exploration of model space,
especially when considering many populations, is crucial for evaluating
robustness of particular features of inferred history.

Finally, natural selection distorts allele frequencies not only at sites
directly under selection, but also in linked regions of the genome through
background selection. This can have a sizeable impact on the SFS and lead to
biases in demographic inference \cite{ewing2016consequences, johri2021impact}.
In gene-sparse species, such as primates, considering data from regions away
from functionally constrained elements is feasible, as we did in example 2
here.  Alternatively, the genome may be partitioned by $B$-value (a
measure of the local strength of background
selection, \cite{mcvicker2009widespread}) and regions experiencing low levels of
background selection may be used (for an example,
see \cite{medina2023demographic}). In some species with gene-dense genomes, such as
\emph{Drosophila} species, this may be infeasible, and interspersed putatively
neutral variation must be used, such synonymous variation or short introns,
which will still experience background selection. The appropriate strategy will
depend on the genome architecture of the species and genomic resources
available.

>>>>

Using the split-with-migration model defined above (illustrated in
Figure~\ref{fig:im}A), we simulated data for 30 diploid individuals from both
populations using \msprime \cite{baumdicker2022efficient}. These simulations
consided of 500 regions, each of length 1 Mb, with per-base recombination and
mutation rates of $10^{-8}$. The joint SFS was computed by using \tskit's
\texttt{ts.allele\_frequency\_spectrum()} function separately on each replicate
region and summing across regions to find the total SFS across 500 Mb of data
(Figure~\ref{fig:im}B--D).

@@

The model that we wished to fit is relatively simple but still has many
parameters ($11$ in the example we present), so we constructed it in stages by
first fitting simpler one- and two-population models.  Throughout, we used
$u=1.5\times 10^{-8}$ as an estimate of the mutation rate per generation and 29
years as the average generation time. When the goal is to infer a complex
demographic history involving multiple populations, we typically start simple,
with subsets of the full data, and sequentially add complexity.  Here, we began
by fitting one-population models with ancestral and recent size changes for MSL
and GBR.  For the MSL model, we allowed three epochs with constant effective
size in each epoch. We attributed to GBR a sharp contraction $\sim$60 kya
followed by exponential growth to represent the out-of-Africa bottleneck and
subsequent rapid expansion. YAML files representing the models and parameters
in this section can be found in the linked GitHub repository.

@@

We next stitched these one-population models together and refitted the 
resulting demographic model to get a two-population model 
that included continuous symmetric 
migration between MSL and GBR and an ancestral size expansion. Following this,
we added a Neandertal branch and fitted the 
Neandertal-modern human divergence time, the Neandertal effective size, and the 
Neandertal-to-GBR admixture proportion. We also fitted the divergence time 
between the sampled Vindija Neandertal lineage and the Neandertal lineage which 
admixed with modern non-African humans. Finding that this parameter tended to 
converge to its lower bound (imposed by the estimated time of death of the 
Vindija Neandertal, ~55 kya), we fixed it to a value supported by literature 
(90 kya, \cite{prufer2017high}) in a second round of optimization.
Rather than fitting it, we fixed the time of the Neandertal-GBR pulse to 50 kya.
Fixing parameters for which good estimates are available in the literature 
(see, e.g., \cite{sumer2025earliest}) is often reasonable in the first stages of 
model optimization, and we expected this value to be especially difficult to fit 
due to the small pulse proportion and Neandertal sample size. In the second 
round of optimization on the three-population model, we refit all parameters 
aside from the two event times mentioned above to refine the estimates.

@@

The site frequency spectrum is just one of many options for inferring
evolutionary parameters from genetic data. It is straightforward to compute
from data, and there are many methods for predicting the SFS, making it an
appealing framework for demographic inference. As shown here in two examples,
we have incorporated the \demes format for specifying demographic models into
\moments to easily compute the SFS and perform demographic inference, including
computing uncertainties.

\moments is limited in the size of the dataset that it can consider. Sample
sizes can be large (in the hundreds), especially when considering models with
one or few populations, but memory issues will develop when sample sizes are
extreme or as the number of populations grows. We are limited to a maximum of
five populations existing at any one time, including ancient samples. Other
SFS-based methods, such as \momi \cite{kamm2020efficiently, dilber2024faster},
can include additional populations, typically with smaller sample sizes.

@@

When inferring models with multiple populations, especially four or more, both
the numbers of parameters and the space of plausible model topologies can grow
rapidly. It can therefore become challenging for the optimization routines to
find optima in parameter space, and overparameterization can become a problem.
One hint that models are overparameterized (or models misspecified) is runaway
behavior of parameter values, in which they converge to unreasonably small or
large values (or approach the imposed bounds). Starting with simple
parameterizations and sequentially adding complexity helps to alleviate this
issue. The challenge of large model space is more difficult to address, and
must be handled through building and testing a wide range of plausible model
topologies. While time-consuming, a broad exploration of model space,
especially when considering many populations, is crucial for evaluating
robustness of particular features of inferred history.

Finally, natural selection distorts allele frequencies not only at sites
directly under selection, but also in linked regions of the genome through
background selection. This can have a sizeable impact on the SFS and lead to
biases in demographic inference \cite{ewing2016consequences, johri2021impact}.
In gene-sparse species, such as primates, considering data from regions away
from functionally constrained elements is feasible, as we did in example 2
here.  Alternatively, the genome may be partitioned by $B$-value (a measure of
the local strength of background selection, \cite{mcvicker2009widespread}) and
regions experiencing low levels of background selection may be used (for an
example, see \cite{medina2023demographic}). In some species with gene-dense
genomes, like \emph{Drosophila} species, this may be infeasible, and
interspersed putatively neutral variation must be used, such as synonymous
variation or short introns, which will still experience background selection.
The appropriate strategy will depend on the genome architecture of the species
and genomic resources available.
